---
title: "Chapter 1 Notes"
author: "John Lockwood"
format: html
editor: visual
---

These are notes are based on *Discovering Statistics Using R*, part of my [Statistics 1](../DiscoveringStatisticsUsingR/Statistics_01.qmd) course.

## Well, here we are

This chapter gets into "Why Statistics", with some entertaining arm-waving about how answering questions in the sciences often involves processing "quantitative methods." There's also a flow chart about the research process, which the rest of the chapter goes through. It starts to get interesting as he develops theories about why game show contestants have narcissistic personality disorder, and shows a data table. A good theory should generate testable — ideally falsifiable — hypotheses.

Moves on to discussing data collection. When we do this we need to figure out:

1\) What to measure.

2\) How to measure it.

### What to Measure

Independent vs dependent variables. If we think one thing causes another, the cause an independent variable. The thing it causes is the dependent variable. Especially use these terms when we use experimental methods, because we can manipulate the variables. More generally, a predictor variable corresponds to independent variable and outcome variable refers to a dependent variable, so we can also use in corelational or cross-sectional research where we don't manipulate variables. These terms don't imply causation as strongly.

Categorical variables (enums, more or less), vs continuous variables (real (or at least natural) numbers). Some categories are binary variables (male/female, alive/dead). Of course, categories can also be represented by numbers.

Ordered categories are ordered (e.g. contest winners could be labeled first, second, third, etc) nominal categories are just names and don't tell us about ordering (Bob, Jane, Phred with a ph).

Continuous variables include interval variables. "To say that data are interval, we must be certain that equal intervals on the scale represent equal differences in the property being measured." E.g. rating scales, 3 vs 4 same as 4 vs 5 (but some self-report scales are more subjective and therefore ordinal? ). Discussion is confusing. Ratio variables are interval variables with the added condition that ratios are meaningful. A professor rated 4 is twice as helpful as one rated 2. Reaction time as a better example, the ratio of 500 ms to 400 ms is the same as 1000:800 ms.

Continuous variables can be measured discretely. E.g. age is measured in years, not in hours or picoseconds.

Measurement errors. Could be from improperly calibrated equipment, differences in measurement equipment, and notoriously true of self-reported data, where people may be interested in protecting their reputation, saying what they expect the researchers want, etc. etc.

Validity vs reliability. Validity: Did the measurement really report what it was designed to measure? Reliablitity: Is it consistent in its measurements.

### Data Collection: How to Measure

#### Correlational Research

AKA cross-sectional research. Where we observe what naturally happens in the world.

Correlational research has problems establishing cause, often because of a **tertium quid** (third thing) that is related to and perhaps causes both what we considered the cause and the effect, or perhaps that causes the effect independently of the cause. These extra factors are also sometimes called **confounding variables** or simply confounds.

#### Experimental Research

Where we manipulate one variable to see its effect on another. Often we're looking to eliminate confounding effects by making sure we don't see an effect absent the cause, so we test with the cause and without it.

### Two methods of data collection:

1.  Called "between-groups", or "between-subjects" or "independent design". We manipulate the independent variable by having two different groups of subjects.
2.  Called "whithin-subjects" or "repeated measures design." E.g. take the same group and give them negative reinforcement for a period of weeks and test performance, and then in a subsequent period(s) give them positive or no reinforcement (or both), and compare the performance.

### Two types of variation

1.  If the same subject performs slightly differently on the same test under the same conditions at different times, this is **unsystematic variation**.
2.  If we change the conditions between one run and the next, then (you guessed it), this is **systematic variation**.

Same things happen in groups of different participants. Given the same test to two groups, one group will do better than the other do to individual differences. This is unsystematic.

But there are always both things going on, in any test, so the role of statistics is to sort out how much is due to individual (or group) differences, and how much do to the manipulated independent variable.

We try to minimize the influence of unsystematic variation by **randomization**. In repeated measure design, you can minimize the effects of boredom (negative) and practice effects (i.e. learning, positive), by changing the order of sessions for different participants (i.e experiment vs control first). This is called **counterbalancing** the order. In independent design, you need to ensure that you select your groups randomly, especially with respect to any confounding variables.

### Frequency distributions

When you have data, it's useful to do a histogram of the frequency of different values. One way these can be distributed is a normal distribution, the good ol' bell curve. Height is a good example.

Two ways to vary from normal distribution are: 1. **Skew**, where values cluster to the left or right of center. Note (this is counter-intuitive) positively skewed means the **tail** points to the right, so the hump is on the left. Negative skew is where the tail points to the left, so hump is in the right. 2. **Kurtosis** has to do with how pointy the distribution is and how many clusters are in the tails. A tall pointy distribution with heavy tails is **leptokurtic**, or has **positive kurtosis**. A more boxy, confused disturbutions with highs throughout, flatter than normal, has negative kurtosis or is **platykurtic**.

### The center of a distribution

We can use three diffferent measures for center.

The mode occurs most frequently, i.e. the tallest bar of our histogram.

(Interlude: Where's my histogram? Come on, Andy, where's my code? I guess we don't get to do R just yet. Talking makes programmers sad. Be patient, grasshopper.)

Mean, mode, and median. Mean is the arithmetic average found in the usual way:

$$\frac{\sum\limits_{1 \to n}}{n}$$ The median is (more or less) a score such that 50% of values are below, and 50% above. Formally, we arrange in ascending order, count the number of scores, add one to the count, and divide by two. This gives the index of the median score. For an odd number, works out great. For an even number, let's say 10 scores

$$\frac{10 +1}{2} = 5.5$$

which is midway between scores 5 and 6 so we average those to scores and that's the median.

The mode is the value that occurs most often

Let's use the example he gives and try our hand at some R code. (You can't keep a good man down — and programmers neither.) Google tells us that Tidyverse simply uses R arrays in the one dimensional case, so here are the values he gives, the mean and the mode.

```{R}
# Since I pirated these values from the text, we'll give them a pirate name
arr <- c(108, 103, 252, 121, 93, 57, 40, 53, 22, 116, 98)

cat("Array length:\n")
print(length(arr))

cat("\nMedian value\n")
median(arr)

cat("\nMean value\n")
round(mean(arr), 2)
```

The results I got were consistent with the book. Since there's no repeating value, mode is undefined, and it turns out R's built-in mode function simply returns the data type of the array, "numeric" in this case. However, a Google search suggested this code doing a descending order sort on a table and selecting the first value:

```{R}
my_array <- c(1, 2, 2, 3, 4, 2, 5)
# Returns the name of the highest-frequency element
names(sort(-table(my_array)))[1] 
```

That returns two, as expected. Google cautions against using this code however for general cases, and we can easily see that it fails to handle the case where the length \>= 2 and the first two values are the same, where mode is undefined. We would also need to consider the trivial one-element case.

June 25, 2026\
We begin today by just doing the self-test, which in R amounts to removing the 252 value and re-running the mean. By way of R pedagogy, let's see if we can do that with respect to arr:

```{r}
# Original arr
arr <- c(108, 103, 252, 121, 93, 57, 40, 53, 22, 116, 98)
arr

# Use a negative index to remove value at position 3
short_arr <- arr[-3]
short_arr

print(round(mean(short_arr), 2))
```

OK, my little excuRsion agreed with the text, cool. The point of this was to show that compared to the median, the mean is sensitive to outliers.

He next shows how to get the **range**, which is just the difference between the min and max value. This is easy enough to find in R through the usual procedure of "guess how to do it and run it to see if it works." (Statements like that is why other senior engineers used to object to me sharing their title back in the days when that's what I was doing).

```{R}
library(reticulate)
arr <- c(108, 103, 252, 121, 93, 57, 40, 53, 22, 116, 98)
short_arr <- arr[-3]

print("Showing the range for the original array")
range_long = range(arr)
print(range_long)
print(diff(range_long))

# For the self test the range is much smaller
print("For the self test the range is much smaller")
range_short = range(short_arr)
print(range_short)
print(diff(range_short))

```

Gemini [pointed out](https://share.gemini.google/YkD6RjeRtWjB) that that code can be improved for large data sets, where it tends to be inefficient, and provided another way to do it.

### Quartile Data

Quartiles are three values that split the range into four equal parts. First, we calcluate the median, which splits the data into two eauql parts. The lower quartile is the median of the lower half, and the upper quartile is the median of the lower half.

Can we accomplish this in R easily? I would imagine so. We can use the quantile function.

```{R}
arr <- c(108, 103, 252, 121, 93, 57, 40, 53, 22, 116, 98)
print(length(arr))
sorted_arr = sort(arr) # Show the sorted array
print(sorted_arr)
print(quantile(arr)) # Use the original, quantile will sort for us

# Next we'll show the inter-quartile range, which Fields discusses next.
IQR(arr)
IQR(arr, type=1) # How to match the book's answer.
```

Note that the values obtained sing quantile are intermediate between actual values, the true medians. Perhaps before I finish up the chapter I'll write a function to subset the array appropriately (R has "cut" function, but I'm not sure that's fit for this purpose).

Along similar lines, I've printed out the IQR twice here. Using type=1 gets an answer that will agree with the textbook, which uses actual values from the array. Using the default uses the quantile function under the hood, which gives interpolated values, (i.e. 57 = 112-55). For background see \[this Gemini response\](<https://share.gemini.google/2xcibEUZqR28>) and the [IQR documentation](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/IQR)

Next, let's do the self-test using R. Again, in the textbook, we haven't gotten to R yet, I'm skipping ahead. The book is aimed at undergrads who haven't learned to program yet, not old guys who used to do it for a living.

The self test asks us to compute several values for an array of values, representing the time in seconds that a group of heavy smokers put on a treadmill fell off from exhaustion. Let's set that up and solve it using R defaults.

```{R}
data <- c(18, 16, 18, 24, 23, 22, 22, 23, 26, 29, 32, 34, 34, 36, 36, 43, 42, 49, 46, 46, 57)


# print(data)

# We're asked to compute the mode, median, mean, upper and lower quartiles, range, and interquartile range.

# Let's first print the sorted data to give us an idea of what's going on.
s_data <- sort(data)
cat("Data:", s_data)

# No value occurs "the most times", though if I had to guess I'd say something between 34 aand 36, which both appear twice, as do 22 and 46
cat("\nNo mode")

# The rest of the tasks are straightforward if we focus on using R rather than matching whatever value Andy publishes (see discussion above.)
cat(sprintf("\nMean: %.2f", mean(data)))
cat(sprintf("\nMedian: %.2f", median(data)))
cat("\nQuartiles:\n")
print(quantile(data))
cat(sprintf("Inter-quartile range: %.2f\n", IQR(data)))
```

In the next section he discusses how probable it is that something could happen, using suicide data to show that most suicides occur between the ages of 30-35 (which isn't borne out here in US by real data, I think). There's more to do in this section, but we'll leave it there for today.

### An aside about Reticulate

June 30, 2026

OK, I have some more of the book to get through for chapter 1, but since I only have a short time today, I thought this might be a good time to try my hand at getting some R data into Python.\
\
Reticulate has an np_array function, so in the tradition of all famous last words, let's see if it's just that easy:

```{r}
library(reticulate)


# Ensure we still have this.
print(data)


# Ensure we can access the R array
np_data = np_array(data, order="F")

# Let's see if we can access a basic numeric value (integer)
my_int <- 42L


```

OK, that seemed to work, it didn't cough or sputter. But the acid test is whether we can use it in Python. Let's just try printing it for now to see if it's reasonable. In quarto, the reticulate package creates an object, "r", through which we can access objects in r's memory

```{python}
import numpy as np
print("Can we access the R array now?")
print(r.np_data)

print("\nAccess the integer we declared in R too.")
print(r.my_int)

print("We'll now try to go the other way to see if that works")

greeting = "Hello from Python!"
```

Can we get our Python data back out into R?

Yes, that works too. Cool beans.

```{r}
print(py$greeting)
```

Clearly our examples at this point are all fairly trivial proofs of concept. For more on reticulate, check out the [package documentation](https://cran.r-project.org/web/packages/reticulate/refman/reticulate.html) or this [RStudio](https://rstudio.github.io/reticulate/index.html) version of the same.

## Doing It In Python

Now that we have our R array moved over to Python, let's see briefly demonstrate how to do what we've done so far using numpy's API.

```{python}

import numpy
import scipy.stats as stats # for mode

data = r.np_data

print("The data from R:")
print(data)

print("\nThis array has", len(data), "elements.")

print("\nThe mean is %.2f." % data.mean())
print("\nThe median is %.2f." % np.median(data))


# Get the mode
mode_array, mode_count = stats.mode(data)
print("\nThe mode array contains %d elements", mode_count)
print("The mode array value is", mode_array)


print("\Here is the quantile (quartile) data:")
quantile_data = np.quantile(data, [0, .25, .5, .75, 1])
print(quantile_data.astype(int))

```

Until now, the functionality of R and Python have pretty much been equivalent. Using scipy, we have an implementation of mode that R seems to lack, but it returns only one of the matching "most frequent" values ([as documented](https://docs.scipy.org/doc/scipy/reference/generated/scipy.stats.mode.html)). Given the complexity of capturing the "mode" idea for the various cases, I'm not sure having an opinionated implementation is better than not having one, but we leave that for the language warriors to sort out.

## Using Frequency Distributions to Estimate Probability ("Go Beyond the Data").

The idea here is that in a normal distribution (where values are centered symetrically around a mean), we can convert an actual score into a standard score or z-score. See also and for the formula below, \[Standard Score\](<https://en.wikipedia.org/wiki/Standard_score>).

A standard score is relative to an idealized case with a mean of zero and a standard deviation of 1. To obtain a z-score, we proceed as follows: We take the actual data point we're interested in, and subtract the mean of all the scores, then we divide that result by the standard deviation.

Or,

$$z = \frac{x - \mu}{\sigma}$$

Where $x$ is our data point, $\mu$ is our population mean, and $\sigma$ is our population stand deviation.

I.e.

The example given in the book is for suicide data by age, and we assume the data has a mean of 36 and a stand deviation of 13. We want to know what the probability of someone 70 or older committing suicide of a given cliff, so we first calculate the z score as follows:

```{r}
z = (70 - 36)/13
round(z, 2)
```

Actually the book didn't show the R code for that, but it was easy enough to figure out.

We can use this as we did in the book's example to look up the probability in Appendix A.1, the table of the standard normal distribution. Thumbing ahead most un-programmatically to to page 934 and looking in the smaller values column, we find the probability to be .00440, or .44%.

Hooray? Can we use a table of logarithms next to multiply numbers? Hello, this is John calling from the 21st century, and we have computers now.

Well, Google relates that we can get a standard z score in R using the scale function, so let's try that, since we looked it up. Unfortunately that isn't exactly the function we used above, it wants to operate on an array or vector or matrix or the like. Let's see if Google can make up a nice vector of about 20 values with a mean of 36 and a standard deviation of 13. (See \[this query\](<https://share.gemini.google/cn0MLW8z28Ny>))

```{r}

# Some test data with a value we want to check
suicide_data = c(10, 22, 24, 26, 27, 27, 28, 29, 30, 33, 39, 41, 42, 43, 44, 44, 45, 47, 49, 70)
suicide_data

# Scale vector and get the last element (formerly 70) into a variable
scaled = scale(suicide_data, 36, TRUE)
scaled_70 = scaled[20]
print(round(scaled_70, 2))
```

Once I'd correctly refined the request, Gemini correctly gave us a set of fake data , also according to a Google query, the pnorm function will allow us to convert this 2.62 back into a probability.

```{r}
scaled_70 = round(scaled_70, 2)
scaled_70
probability = pnorm(scaled_70, lower.tail = FALSE)
probability
cat(sprintf("Suicide probability age 70+: %0.2f%%.", probability * 100))
```

In retrospect, we could have used pnorm to approximate this answer in a virtual one-liner, by providing our mean and our standard deviation to scale automatically:

```{r}
suicide_data = c(10, 22, 24, 26, 27, 27, 28, 29, 30, 33, 39, 41, 42, 43, 44, 44, 45, 47, 49, 70)

probability = pnorm(suicide_data, mean=36, sd=13, lower.tail = FALSE)[20]
print(probability)
cat(sprintf("Suicide probability age 70+: %0.2f%%.", probability * 100))
```

It's interesting that the value we get is not the same as we get if we scale first, though it's fairly close.

Given the fun we're having climbing the rest of the steep part of the learning curve, we won't stop to explore that mystery further.
